What is an Ancestral Recombination Graph?

### 2022-02-14 paper.tex
Trivial grammar corrections
More text

The Griffiths ARG encoding is ambiguous because it does not explicitly
define which of the parent lineages resulting from a recombination
event carries ancestral material to the left or right of the
breakpoint. The ambiguity is resolved by adopting a simple rule when
traversing the graph in order to generate the marginal trees for a
given position $z$: at a recombination node $u$, we follow the
first edge in which $u$ is a child if $f(u) \leq z$, and the
second edge otherwise. This ordering requirement, which straightforward
% Saying "awkward" too many times here, rephrase.
to describe can be awkward in practise. For example,
simulating an ARG in this representation is considerably
complicated by the fact that the
first event to occur may hit the right-hand lineage rather
than the left.
Such awkwardnesses can be worked around, of course,
but in practise methods such has
\texttt{beagle}~\citep{lyngso2005minimum} and
\texttt{KwARG}~\citep{ignatieva2021kwarg}
resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event.

@@

The limitations of the Griffiths ARG encoding are a direct consequence
of its derivation from the coalescent with recombination stochastic process.
Indeed, these limitations arise because it is such a \emph{good}
mathematical encoding, providing sufficient information to
define the realised ancestry under the assumptions of the model,
and no more.

We describe here an alternative formulation in which the
concepts are based on the diploid pedigree, which
is free from the limitations discussed in the previous section.
\citet{mathieson2020ancestry} and others
\citep{wakeley2012genetics,gusfield2014recombinatorics,speed2015naturereviewsgenetics}
have also suggested this approach, although without explicitly defining
the roles of nodes and the details of how ancestry
flows along edges.
\begin{figure}
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.3\textheight]{illustrations/pedigree}
    \caption{Pedigree with genetic inheritance}
    \label{fig:pedigree}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.3\textheight]{illustrations/pedigree_ARG}
    \caption{Graph representation of genetic inheritance}
    \label{fig:ARG-of-pedigree}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.3\textheight]{illustrations/pedigree_trees}
    \caption{Tree representation of genetic inheritance}
    \label{fig:ts-of-pedigree}
  \end{subfigure}
\caption{\label{fig-pedigree-and-arg}
The (unannotated) ARG is closely related to the pedigree
graph, where each ``individual'' node is replaced by two ``genome''
nodes, and edges between parents and children are only included if there is
transfer of ancestral material. We usually only include nodes in an ARG
that affect our sample, so that, for example, individuals that are ancestral
to only one sampled genome are not included.}
\end{figure}

[Here are the basic components of the model]

\begin{enumerate}
\item A node corresponds to one of a diploid individual's two genomes.
This is linked explicitly to the pedigree, and is fully general.
Nodes \emph{may} be labelled with certain flags describing their properties,
but it's not necessary.
\item An edge between two nodes defines a genetic inheritance relationship
between two nodes. Edges are annotated with information about the
passage of ancestral materal. There are different forms these annotations
can take. For example, we can follow the approach taken in the Griffiths
ARG and just annotate recombinant edges with the interval $[0, x)$ or $[x, L)$,
depending on side of the breakpoint the edge is on. In the most
general case, we annotate each edge with the full set of disjoint
intervals of ancestral material that flows along that edge.
\citep{mcgill2013graphml} also proposed full ancestral material being
associated with each edge.
\end{enumerate}

\begin{figure}
\vspace{5em}
% NB: not clear if this figure is better off on its own or should be
% merged with the pedigree one.
\caption{\label{fig-arg-annotations}
The specific path through the ARG for a given genome position (and therefore
the local genealogical tree) cannot be determined without annotations
to the graph. (A) The classical Griffiths approach annotates recombination
nodes with the corresponding breakpoint. (B) We can equivalently
annotate the \emph{edges} with the genomic intervals carrying ancestral
material. Edge annotations are somewhat more general, and have
significant computational advantages.}
\end{figure}

\section*{Relationship with other types of phylogenetic networks}

@@

Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
A reticulation loop that does not share any node with another reticulation loop is termed a gall
\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
and the tractability afforded by their relatively simple structure results in a number of attractive
combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
an ARG, without applying any of these restrictions.
% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.

@@

\begin{itemize}
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\item Interchange is a real problem, and if we're going to realise the
potential for ARGs in genomics we must solve it. Efforts to standardise
phylogenetic networks haven't done very well (~\citep{cardona2008extended}
has 89 citations). The only real effort to standardise an ARG format
based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
GraphML~\citep{mcgill2013graphml} has some of the same key ideas
as tskit though, in particular allowing the annotation of ``live sites''
to an edge as a set of disjoint intervals that hold ancestral material.

>>>>

The Griffiths ARG encoding is ambiguous because it does not explicitly
define which of the parent lineages resulting from a recombination
event carries ancestral material to the left or right of the
breakpoint. The ambiguity is resolved by adopting a simple rule when
traversing the graph in order to generate the marginal trees for a
given position $z$: at a recombination node $u$, we follow the
first edge in which $u$ is a child if $f(u) \leq z$, and the
second edge otherwise. This ordering requirement, which straightforward
% Saying "awkward" too many times here, rephrase.
to describe can be awkward in practice. For example,
simulating an ARG in this representation is considerably
complicated by the fact that the
first event to occur may hit the right-hand lineage rather
than the left.
Such awkwardnesses can be worked around, of course,
but in practice methods such has
\texttt{beagle}~\citep{lyngso2005minimum} and
\texttt{KwARG}~\citep{ignatieva2021kwarg}
resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event.

@@

The deep connection between the ARG and the pedigree is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions of the ARG have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of the ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the expressivity of the structure.
Following \cite{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
\caption{\label{fig-pedigree-and-arg}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined as arbitrary inheritance
of genomes between diploid (or any ploidy; see text) individuals.
(A) Each individual consists of two genomes, and each genome is
inherited as a (potentially) recombinant mixture of the individual's parent's
genomes. (B) Genomes are modelled as nodes in a graph, and the edges
denote inheritance. Each edge is annotated with genome coordinates
defining the specific region(s) inherited. An ARG is a retrospective
structure defined with
respect to some set of sample genomes (A--D here), and so only
genomes directly ancestral to those samples are included.
(C) This description of inheritance fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume diploidy here for concreteness, but there
is no inherent limitation on ploidy).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inherited directly from $I_4$'s genome $G$ without
recombination.

The focus on modelling genomes rather than events has some interesting
and important consequences. Firstly, the graph may contain nodes that
do not correspond to any particular event in the ancestry of the
samples. For example, node $H$ is the direct ancestor of $D$ in
Fig~\ref{fig-pedigree-and-arg}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that $D$ descends directly from $K$ (see section XXX)
but there is no necessity for this from a representational perspective.
This ability can be useful in practice and could be required if,
% Do we need to spell this out?
for example, we have a deeply sequenced pedigree of
haplodiploids. % e.g. Honeybees? Be great to cite an example!

Secondly, the way that we model recombination is subtly different.
Rather than having a single node that represents the \emph{event},
there are three nodes (representing the recombinant offspring and
two parental genomes) and two edges with associated genome coordinate
intervals (representing the precise inheritance patterns).
For example, in Fig~\ref{fig-pedigree-and-arg}, node $A$ is the
recombinant descendent of $E$ and $F$, inheriting positions
FIXME from $E$ and genome positions FIXME from $F$. Any particular
node can be both a recombinant (it is the recombined product of
its parent's genomes) and a common ancestor (the ancestry from
multiple samples meet at this node; see the next section).
[More stuff here: clarify why we assign coordinates to edges
rather than nodes. Basically lay out all the stuff that we can
talk about \emph{without} going in to the details of
ancestry resolution.]

\textcolor{red}{TODO: reduce this down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature. I think pulling apart the
idea of a ``reticulation'' vs what we call a recombinant
is essential here.}

@@

Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
A reticulation loop that does not share any node with another reticulation loop is termed a gall
\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
and the tractability afforded by their relatively simple structure results in a number of attractive
combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
an ARG, without applying any of these restrictions.

% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.

\section*{Ancestry resolution and simplification}

\begin{figure}
\centering
\vspace{5em}
\caption{\label{fig-ancestry-resolution}
}
\end{figure}

@@

\begin{itemize}
\item It is unfortunate that we are adding yet another meaning to the
term ARG here, but all we're really doing is making the implicit
assumptions that others have made (i.e., describing the output of
Relate, tsinfer etc as ARGs) and making them concrete. We need to have a
term to describe all of these things, and since people are already using
ARG for this, we might as well make it rigorous. We can then move away
from the unfruitful discussion about which methods infer a ``true'' or
``real'' ARG (none do, really, under the original definitions) and
instead start discussing the \emph{properties} of these ARGs. (See
next point also.)
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\item Interchange is a real problem, and if we're going to realise the
potential for ARGs in genomics we must solve it. Efforts to standardise
phylogenetic networks haven't done very well (~\citep{cardona2008extended}
has 89 citations). The only real effort to standardise an ARG format
based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
GraphML~\citep{mcgill2013graphml} has some of the same key ideas
as tskit though, in particular allowing the annotation of ``live sites''
to an edge as a set of disjoint intervals that hold ancestral material.

\item There are real computational disadvantages to the Griffiths ARG
data structure. Having two types of node (three really, as we need
``sample'' nodes which are also treated differently) complicates algorithms,
and makes it difficult to reason about the structure of the ARG as a
graph. For example, there's some weirdness about mutations over
recombination nodes discussed by Gusfield [Ana?]
